Using Hybrid Machine Learning for Fundamental Stellar Parameter Inference

### 2024-05-03 paper/paper2_neo/paper.tex
Update paper
- Add abstract
- Add reconstruction profiling
- Add style thingies

\documentclass[twocolumn]{aastex631}
\usepackage{graphicx}
\usepackage{physics}
\usepackage[super]{nth}
\usepackage{enumitem}

@@

\begin{abstract}
    Text
\end{abstract}
\keywords{Text}

@@

\begin{table}[h!]
    \hspace*{-1.35cm}\begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\textbf{PHOENIX Grid Subset}}\\
        \hline\hline
        Parameter & Symbol & Range\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0] dex\\
        Iron Abundance & [Fe/H] & [-0.5, 0] dex\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K\\
        Surface Gravity & $\log(g)$ & [2, 6]\\
        Wavelength & $\lambda$ & [8038, 12849] \AA\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study. These limits 
    were imposed to reduce the computational cost of the algorithms and to 
    ensure a rectilinear parameter space in order to work with \texttt{scipy}'s
    \texttt{RegularGridInterpolator}. The wavelength limits in particular 
    roughly line up with that of the Habitable Zone Planet Finder (HPF) 
    spectrograph. This subset is comprised of 1314 individual spectra.}
\end{table}

@@

\subsection{Line Identification with \texttt{blase}}
The next step was to convert the PHOENIX subset into an interpretable format.
We wanted to represent the spectra as a list of spectral lines rather than a
list of fluxes. This was done using \texttt{blase}, which detects spectral 
lines as Voigt profiles and tunes the profiles to mimic the original 
PHOENIX spectrum with back propagation. Four parameters were optimized: the 
line center $\mu$, the log-amplitude $\ln(a)$, the Gaussian width $\sigma$, and the 
Lorentzian width $\gamma$. The optimization used the Adam optimizer with a
learning rate of 0.05 over 100 epochs. In addition, we limited two custom 
parameters: wing cut to 6000 and prominence to 0.005. Wing cut is a parameter
that determines the extent of the Voigt profile to evaluate, saving
computational resources by not evaluating small numbers. Prominence sets a
lower limit for the amplitude of detected lines, which also saves resources 
by disregarding shallow lines. In short, larger values for wing cut and 
smaller values for prominence both increase the accuracy of using
\texttt{blase}'s line detection and optimization at the expense of an
increased computational cost. In addition, it should be mentioned that
\texttt{blase} uses the pseudo-Voigt approximation, which saves on 
computational cost while remaining accurate to about 1\%. It is a 
weighted average of a Gaussian and Lorentzian as opposed to a convolution.
\texttt{blase}'s pseudo-Voight profile implementation uses the following:
\begin{gather}
    \tilde{\mathsf{V}}_\mu(\lambda) = \eta L(\lambda - \mu'; f) + (1 - \eta)G(\lambda - \mu'; f)\\
    \nonumber \eta = \sum_{n=1}^3 a_n\left(\dfrac{2\gamma}{f}\right)^n \qquad a = \mqty[1.36603 \\ -0.47719 \\ 0.11116]\\
    \nonumber f = 32\sum_{n=0}^5 b_n \left(\sqrt{2\ln(2)}\sigma\right)^{5-n}(\gamma)^n \qquad b = \mqty[1 \\ 2.69269 \\ 2.42843 \\ 4.47163 \\ 0.07842 \\ 1]
\end{gather}
Notice that we use $\mu'$ instead of $\mu$ in the formula. This is because
\texttt{blase} allows the line center to shift slightly during optimization, 
and it is this shifted center which is used in computation. The individual 
Voigt profiles are still indexed by $\mu$ for cross-model line identification,
explained in the next section. Once optimization was complete, the list of 
identified lines, present in the state dictionary of \texttt{blase}'s model, 
was saved to a \texttt{.pt} file for each PHOENIX subset grid point. The total 
disk space these files took up is 465 MB.

@@

\begin{figure*}[t!]
    \centering
    \includegraphics[width=\textwidth]{images/figure1.png}
    \caption{The number of detected spectral lines at each grid point of 
    a slice of the PHOENIX subset at solar metallicity. We can see that
    the number of detected lines decreases with increasing $T_{\mathrm{eff}}$.
    Also note that from $T_{\mathrm{eff}} = 7000$ K onward, the spacing 
    between grid points increases from 100 K to 200 K.}
\end{figure*}

@@

\begin{figure*}[t!]
    \centering
    \includegraphics[width=\textwidth]{images/figure2.png}
    \caption{Heatmap showing how $\ln(a)$ and $\sigma$ vary over the PHOENIX subset
    slice at solar metallicity. Notice the missing chunk in the top left
    of the figure; \texttt{blase} did not detect a spectral line here,
    but we have to artificially populate those points with lines that 
    have $\ln(a) = -1000$. The change in spacing on the x-axis is owing to
    the step of $T_\mathrm{eff}$ in the PHOENIX grid changing from 100 K to 
    200 K after $T_\mathrm{eff} = 7000$ K.}
\end{figure*}

@@

\begin{figure*}[t!]
    \centering
    \includegraphics[width=\textwidth]{images/figure3.png}
    \caption{Heatmap showing how $\ln(a)$ and $\sigma$ vary over the PHOENIX 
    subset slice at solar metallicity, now supersampled with the PHOENIX 
    generator. Notice that the missing chunk in the top left still exists 
    and does not display any artifacts, as the artificially populated points 
    are removed after interpolation to retain the model's integrity. Also
    see that the x-axis spacing is now uniform, as the PHOENIX generator 
    was evaluated at constant step.}
\end{figure*}

\section{Bayesian Inference and Testing}
\subsection{Inference Algorithm}
With the PHOENIX generator complete, we used Bayesian optimization as the
inference algorithm, specifically the \texttt{gp\_minimize} function from
the \texttt{scikit-optimize} library. This algorithm uses a Gaussian Process
to model the objective function, which in this case was the RMS 
(Root-Mean-Square) loss between the interpolated spectrum and the true 
spectrum. The optimizer was configured to first run 30 random evaluations to 
seed the surrogate model, then run 20 more evaluations now guided by the 
surrogate model. This totals to 50 evaluations, which was deemed sufficient 
for this study. We then decided to use ensemble methods, running 32 parallel
optimizers to get a more statistically sound result. One inference run 
takes an average of 15 minutes to complete on Triton (see Section 5 for
specifications).

\subsection{Test Data}
To test the performance of the inference algorithm, we used the PHOENIX subset
itself. At first glance, this may seem like a useless endeavor, as the 
PHOENIX generator has indeed memorized the PHOENIX subset, being able to
reconstruct a PHOENIX spectrum when evaluated at that grid point. However,
that is precisely what allows us to use the PHOENIX subset as test data.
The Bayesian optimizer's surrogate model is seeded by random continuously 
sampled generator evaluations within the search space, \textit{not} grid 
points of the PHOENIX subset, meaning the surrogate model has no `memorization' 
to speak of. If the optimizer had been an iterative grid refinement search, 
this would not have been possible, because grid methods rely on just that, 
the grid.

\subsection{Spectral Reconstruction Time}
The typical use case for the PHOENIX generator may be to batch 
reconstruct spectra from an array of input coordinates. Therefore, there
is some motivation to attempt to reduce the computational cost of this
procedure to tractable levels. We evaluate the computational time needed
to use the PHOENIX generator in two distinct ways. First, for a single
input, which would be relevant in serial applications. Second, for an array
of multiple inputs, which would be relevant in parallel applications.
The \texttt{RegularGridInterpolator} API allows for the passing in of 
an entire array of input coordinates to be evaluated at once, however 
using \texttt{blase} to reconstruct the spectrum from our interpolated
state dictionary is done serially, leading to what is actually more of 
a psuedo-parallel evaluation.

\subsection{Bayesian Optimizer Accuracy}
TACC

\section{Discussion}
The computer used for this study, Triton, has the following specifications:
\begin{table}[h!]
    \hspace*{0.6cm}\begin{tabular}{|c|c|}
        \hline
        \multicolumn{2}{|c|}{\textbf{Triton Specs}}\\
        \hline\hline
        CPU & AMD EPYC 7513\\
        Memory & 256 GB\\
        GPU & Nvidia A100 40GB ($\times$2)\\
        \hline
    \end{tabular}
    \caption{This computer is owned by Caroline Morley and is available to
    members of her research group. It was used for all computations, but not 
    for generating visualizations. The EPYC 7513 is a 32c/64t CPU with a 
    boost clock of 3.65 GHz.}
\end{table}

@@

    abundances.

@@

advent of physics-informed machine learning.

\section*{Acknowledgements}
Text

>>>>

\documentclass[twocolumn]{aastex631}
\usepackage{graphicx}
\usepackage{physics}
\usepackage[super]{nth}
\usepackage{enumitem}
\usepackage{natbib}

\graphicspath{{./images/}}

@@

\shortauthors{Shankar \& Gully-Santiago \& Morley}

\begin{abstract}
    The advent of machine learning methods has revolutionized 
    numerous scientific methods, among which is spectroscopic 
    inference. Studies can and have shown that machine learning 
    methods prove effective in inferring what we call
    fundamental stellar parameters. These include $T_{\mathrm{eff}}$,
    $\log(g)$, and [Fe/H] for the purposes of this study but can 
    also encompass other parameters such as [$\alpha$/Fe], C/O, and
    $f_{sed}$. While current methods tend to treat spectra purely
    from a data perspective, we propose to use a hybrid machine 
    learning (HML) method that uses the semi-empirical approach 
    of treating spectra as a set of spectral lines put forth by 
    \texttt{blase}, combined with traditional linear interpolation 
    as well as Bayesian optimization. We interpolate the four 
    line properties ($\mu'$, $\ln(a)$, $\sigma$, $\gamma$) of 
    128,723 unique spectral lines across a subset of the PHOENIX
    synthetic model grid to create the PHOENIX generator, allowing
    us to reconstruct spectra at any continuously valued point 
    lying within the domain of the PHOENIX subset, and even reconstruct 
    batches of spectra in pseudo-parallel fashion, with the speedup 
    exceeding 20 times once more than 60 spectra are reconstructed. 
    We then combine the PHOENIX generator with Bayesian optimization to 
    efficiently infer fundamental stellar parameters. This study is 
    however a proof of concept, as we have not yet included various
    features that would be necessary for a full-fledged tool for 
    spectroscopic inference, however we show that semi-empirical 
    HML methods are a viable, albeit fledgling alternative to 
    more traditional approaches.
\end{abstract}

@@

\begin{table}[h!]
    \centering
    \begin{tabular}{lll}
        \hline
        \bf{Parameter} & \bf{Symbol} & \bf{Range}\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0] dex\\
        Iron Abundance & [Fe/H] & [-0.5, 0] dex\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K\\
        Surface Gravity & $\log(g)$ & [2, 6]\\
        Wavelength & $\lambda$ & [8038, 12849] \AA\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study. These limits 
    were imposed to reduce the computational cost of the algorithms and to 
    ensure a rectilinear parameter space in order to work with \texttt{scipy}'s
    \texttt{RegularGridInterpolator}. The wavelength limits in particular 
    roughly line up with that of the Habitable Zone Planet Finder (HPF) 
    spectrograph. This subset is comprised of 1314 individual spectra.}
\end{table}

@@

\subsection{Line Identification with \texttt{blase}}
The next step was to convert the PHOENIX subset into an interpretable format.
We wanted to represent the spectra as a list of spectral lines rather than a
list of fluxes. This was done using \texttt{blase}, which detects spectral 
lines as Voigt profiles and tunes the profiles to mimic the original 
PHOENIX spectrum with back propagation. Four parameters were optimized: the 
line center $\mu$, the log-amplitude $\ln(a)$, the Gaussian width $\sigma$, and the 
Lorentzian width $\gamma$. The optimization used the Adam optimizer with a
learning rate of 0.05 over 100 epochs. In addition, we limited two custom 
parameters: wing cut to 6000 and prominence to 0.005. Wing cut is a parameter
that determines the extent of the Voigt profile to evaluate, saving
computational resources by not evaluating small numbers. Prominence sets a
lower limit for the amplitude of detected lines, which also saves resources 
by disregarding shallow lines. In short, larger values for wing cut and 
smaller values for prominence both increase the accuracy of using
\texttt{blase}'s line detection and optimization at the expense of an
increased computational cost. In addition, it should be mentioned that
\texttt{blase} uses the pseudo-Voigt approximation, which saves on 
computational cost while remaining accurate to about 1\%. It is a 
weighted average of a Gaussian and Lorentzian as opposed to a convolution.
\texttt{blase}'s pseudo-Voight profile implementation uses the following:
\begin{subequations}
\begin{gather}
    \tilde{\mathsf{V}}_\mu(\lambda) = a\left[\eta \mathbf{L}(\lambda - \mu'; f) + (1 - \eta)\mathbf{G}(\lambda - \mu'; f)\right]\\
    \eta = \sum_{n=1}^3 \mathbf{u}_n\left(\dfrac{2\gamma}{f}\right)^n\\
    f = 32\sum_{n=0}^5 \mathbf{v}_n \left(\sqrt{2\ln(2)}\sigma\right)^{5-n}(\gamma)^n\\
    \nonumber \mathbf{u} = \mqty[1.36603 \\ -0.47719 \\ 0.11116] \quad \mathbf{v} = \mqty[1 \\ 2.69269 \\ 2.42843 \\ 4.47163 \\ 0.07842 \\ 1]
\end{gather}
\end{subequations}
where $\mathbf{L}$ and $\mathbf{G}$ are abbreviations for Lorentzian and 
Gaussian profiles, respectively. Notice that we use $\mu'$ instead of $\mu$ 
in the formula. This is because \texttt{blase} allows the line center to shift 
slightly during optimization, and it is this shifted center which is used in 
computation. The individual Voigt profiles are still indexed by $\mu$ for 
cross-model line identification, explained in the next section. Once optimization 
was complete, the list of identified lines, present in the state dictionary of 
\texttt{blase}'s model, was saved to a \texttt{.pt} file for each PHOENIX subset 
grid point. The total disk space these files took up is 465 MB.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure1}
    \caption{The number of detected spectral lines at each grid point of 
    a slice of the PHOENIX subset at solar metallicity. We can see that
    the number of detected lines decreases with increasing $T_{\mathrm{eff}}$.
    Also note that from $T_{\mathrm{eff}} = 7000$ K onward, the spacing 
    between grid points increases from 100 K to 200 K.}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure2}
    \caption{Heatmap showing how $\ln(a)$ and $\sigma$ vary over the PHOENIX subset
    slice at solar metallicity. Notice the missing chunk in the top left
    of the figure; \texttt{blase} did not detect a spectral line here,
    but we have to artificially populate those points with lines that 
    have $\ln(a) = -1000$. The change in spacing on the x-axis is owing to
    the step of $T_\mathrm{eff}$ in the PHOENIX grid changing from 100 K to 
    200 K after $T_\mathrm{eff} = 7000$ K.}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure3}
    \caption{Heatmap showing how $\ln(a)$ and $\sigma$ vary over the PHOENIX 
    subset slice at solar metallicity, now supersampled with the PHOENIX 
    generator. Notice that the missing chunk in the top left still exists 
    and does not display any artifacts, as the artificially populated points 
    are removed after interpolation to retain the model's integrity. Also
    see that the x-axis spacing is now uniform, as the PHOENIX generator 
    was evaluated at constant step.}
\end{figure*}

\section{Bayesian Inference and Testing}
\subsection{Spectral Reconstruction Time}
The typical use case for the PHOENIX generator may be to batch 
reconstruct spectra from an array of input coordinates. Therefore, there
is some motivation to attempt to reduce the computational cost of this
procedure to tractable levels. We evaluate the computational time needed
to use the PHOENIX generator in two distinct ways. First, for a single
input, which would be relevant in serial applications. Second, for an array
of multiple inputs, which would be relevant in parallel applications.
The \texttt{RegularGridInterpolator} API allows for the passing in of 
an entire array of input coordinates to be evaluated at once, however 
using \texttt{blase} to reconstruct the spectrum from our interpolated
state dictionary is done serially, leading to what is actually more of 
a psuedo-parallel evaluation. Performance results are shown in Figure 4,
and we can see that the pseudo-parallel implementation is much faster.

\subsection{Inference Algorithm}
We elected to use Bayesian optimization as the inference algorithm, specifically 
the \texttt{gp\_minimize} function from the \texttt{scikit-optimize} library. 
This algorithm uses a Gaussian Process to model the objective function, which in 
this case was the RMS (Root-Mean-Square) loss between the interpolated spectrum 
$\mathsf{M}$ and the true spectrum $\mathsf{D}$, defined as:
\begin{gather}
    \mathcal{L} = \ev{(\mathsf{M} - \mathsf{D})^{\circ 2}}^{1/2}
\end{gather}
The optimizer was configured to first run 30 random 
evaluations to seed the surrogate model, then run 20 more evaluations now guided 
by the surrogate model. This totals to 50 evaluations, which was deemed 
sufficient for this study. One inference run took us an average of 15 minutes to 
complete.

\subsection{Bayesian Optimizer Performance}
To test the performance of the inference algorithm, we used the PHOENIX subset
itself. At first glance, this may seem like a useless endeavor, as the 
PHOENIX generator has indeed memorized the PHOENIX subset, being able to
reconstruct a PHOENIX spectrum when evaluated at that grid point. However,
that is precisely what allows us to use the PHOENIX subset as test data.
The Bayesian optimizer's surrogate model is seeded by random continuously 
sampled generator evaluations within the search space, \textit{not} grid 
points of the PHOENIX subset, meaning the surrogate model has no `memorization' 
to speak of. If the optimizer had been an iterative grid refinement search, 
this would not have been possible, because grid methods rely on just that, 
the grid.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figure4}
    \caption{Line chart showing the time taken to reconstruct varying
    numbers of spectra using the PHOENIX generator (lower is better). 
    We can see that the time taken per spectra for the serial 
    implementation hovers around 15 seconds within run-to-run variance, 
    while the pseudo-parallel implementation continually decreases in time 
    taken per spectrum as the number of inputs increases. The speedup factor 
    (higher is better) increases as more spectra are generated, which is 
    also a desirable outcome.}
\end{figure}

\section{Final Considerations}
The computer used for this study, Triton, has the following specifications:
\begin{table}[h!]
    \centering
    \begin{tabular}{ll}
        \hline
        CPU & AMD EPYC 7513\\
        RAM & 256 GB\\
        GPU & Nvidia A100 40GB ($\times$2)\\
        \hline
    \end{tabular}
    \caption{This computer is owned by Caroline Morley and is available to
    members of her research group. It was used for all computations, but not 
    for generating visualizations. The EPYC 7513 is a 32c/64t CPU with a 
    boost clock of 3.65 GHz. The A100 has 6912 CUDA cores.}
\end{table}

This study combines backpropagation, interpolation, and Bayesian optimization
to develop a HML tool for stellar parameter inference. A summative overview 
of the end-to-end process is shown in Figure 5.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure5}
    \caption{A simple diagram showing the essence of the process used in 
    this paper.}
\end{figure*}

@@

    abundances. So while readers may be tempted to try to identify certain 
    spectral lines such as the one we use for our figures, this is not 
    the point. \texttt{blase} is \textit{agnostic} to the identity of the line
    that it is optimizing. We study these lines as \texttt{blase} sees them
    (i.e. their four shape parameters), because for the purposes of this study, 
    that is the only information that is useful.

@@

advent of physics-informed machine learning \citep{test} \citep{test2}.

\begin{acknowledgements}
    Text
\end{acknowledgements}

\software{Text}

\begin{thebibliography}{dummy}
    \bibitem[Test 2024()]{test} Woop
    \bibitem[Test2 2024()]{test2} Woop woop
\end{thebibliography}
